---
title: "Hands-on Exercise 2: Beyond ggplot2 Fundamentals"
date: "10 January 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
editor: visual
---

## Overview

In this hands on exercise, several ggplot2 extensions will be tried to create more elegant and effective graphics.

-   control the placement of texts on graph utilising [ggrepel](https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html) package

-   create professional publication quality figure utilising [ggthemes](https://ggplot2.tidyverse.org/reference/ggtheme.html) and [hrbrthemes](https://www.rdocumentation.org/packages/hrbrthemes/versions/0.1.0) package

-   plot composite figure utilising ggplot2 and [patchwork](https://patchwork.data-imaginist.com) package

## Getting started

### Importing the required libraries

there are 4 new libraries that are going to be utilised in this hands on. They are:

1.  ggrepel: to repel overlapping text
2.  ggthemes: to provide wider themes, geoms, and scales for ggplot2
3.  hrbrthemes: to provide typography-centric themes and theme componens for ggplot2
4.  patchwork: to combine figures

```{r}
pacman::p_load(ggrepel, patchwork, ggthemes, hrbrthemes, tidyverse)
```

### Importing the data

the data that will be used in this hands-on is exam data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

preview 5 row of the exam data

```{r}
head(exam_data,5)
```

## Beyond ggplot2 annotation: ggrepel

First, we look on the plot without using the ggrepel

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              size=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

## The code

```{r}
#| eval: false
ggplot(data = exam_data,
       aes(x = MATHS,
                 y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              size = 0.5) +
  geom_label(aes(label=ID),
             hjust = .5,
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores")
  
```
:::

There are too much annotation that you cannot see anything anymore

### working with ggrepel

::: panel-tabset
## The plot

```{r}
#| echo: false
ggplot(data=exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              size = 0.5) +
  geom_label_repel(aes(label = ID),
                   fontface = "bold") +
  coord_cartesian(xlim = c(0,100),
                  ylim = c(0,100)) +
  ggtitle("English scores versus Math schores for Primary 3")
```

## The code

```{r}
#| eval: false
ggplot(data=exam_data,
       aes(x = MATHS,
           y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              size = 0.5) +
  geom_label_repel(aes(label = ID),
                   fontface = "bold") +
  coord_cartesian(xlim = c(0,100),
                  ylim = c(0,100)) +
  ggtitle("English scores versus Math schores for Primary 3")
```
:::

now the annotation are not overlapping and only shown when possible

## Beyond ggplot2 themes

there are some built in themes from ggplot such as `theme_gray(), theme_bw(), theme_minimal()` and some more.

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = "grey25",
                 fill = "grey90") +
  theme_gray() +
  ggtitle("Distribution of Math scores")
```

### ggtheme

ggtheme is part of ggplot. However, it is open for extension therefore there are more ggplot that created by communities, `theme_economist()` is one of the example of theme created by others

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = 'grey25',
                 fill = 'grey90') +
  ggtitle('Distribution of Math scores') +
  theme_economist()
```

### hrbrthemes

hrbrthemes are more focused on the typography instead of the overall beauty

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = 'grey25',
                 fill = 'grey90') +
  ggtitle('Distribution of Math scores') +
  theme_ipsum()
```

```{r}
ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = 'grey25',
                 fill = 'grey90') +
  ggtitle('Distribution of Math scores') +
  theme_ipsum(axis_title_size = 18,
              base_size = 15,
              grid = "Y")
```

## Beyond Single Graphs

sometimes plotting a single graph give minimum information. therefore, it is common to plot more than 1 graphic at the same time. this is to increase the amount of information and understanding extracted from the plots.

```{r}
p1 <- ggplot(data = exam_data, 
             aes(x = MATHS)) +
  geom_histogram(bins = 20, 
                 boundary = 100,
                 color = "grey25", 
                 fill = "grey90") + 
  coord_cartesian(xlim = c(0,100)) +
  ggtitle("Distribution of Math scores")

p1
```

```{r}
p2 <- ggplot(data = exam_data, 
             aes(x = ENGLISH)) +
  geom_histogram(bins = 20, 
                 boundary = 100,
                 color = "grey25", 
                 fill = "grey90") +
  coord_cartesian(xlim = c(0,100)) +
  ggtitle("Distribution of English scores")

p2
```

```{r}
p3 <- ggplot(data = exam_data, 
             aes(x = MATHS,
                 y = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              size = 0.5) +
  coord_cartesian(xlim = c(0,100),
                  ylim = c(0,100)) +
  ggtitle("English scores versus Math scores for Primary 3")

p3
```

here are 3 graphs plotted one after another. this way of plotting makes it hard to compare between them

### Creating Composite Graphics: patchwork methods

patchwork is an extension of ggplot2

-   \+ or \| is used to create new column layout

-   () is used for subplots

-   / is used for new row layout

### 2 Plots

::: panel-tabset
## column

```{r}
p1 + p2
```

## row

```{r}
p1 / p2
```
:::

### 3 Plots

::: panel-tabset
## column

```{r}
p1 + p2 + p3
```

## row

```{r}
p1 / p2 / p3
```

## combination

use "\|" instead of "+" to use combination

```{r}
(p1/p2)|(p3)
```
:::

### annotation(tag)

```{r}
((p1/p2)|p3) +
  plot_annotation(tag_levels='I')
```

### plot inside of plot

```{r}
p3 + inset_element(p2, 
                   left = 0.02, 
                   bottom = 0.7, 
                   right = 0.5, 
                   top = 1)
```

### patchwork + ggthemes

```{r}
patchwork <- (p1 / p2) | p3
patchwork & theme_economist()
```

## Reference

-   [Patchwork R package goes nerd viral](https://www.littlemissdata.com/blog/patchwork)

-   [ggrepel](https://ggrepel.slowkow.com/)

-   [ggthemes](https://ggplot2.tidyverse.org/reference/ggtheme.html)

-   [hrbrthemes](https://cinc.rud.is/web/packages/hrbrthemes/)

-   [ggplot tips: Arranging plots](https://albert-rapp.de/post/2021-10-28-extend-plot-variety/)

-   [ggplot2 Theme Elements Demonstration](https://henrywang.nl/ggplot2-theme-elements-demonstration/)

-   [ggplot2 Theme Elements Reference Sheet](https://isabella-b.com/blog/ggplot2-theme-elements-reference/)
