---
title: "Hands-on_Exercise 4b: Visual Statistical Analysis"
date: "28 January 2024"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
editor: visual
---

## Overview

in this hands on exercise, we we explore:

-   ggstatsplot package to create visual graphics with rich statistical information

-   performance package to visualise model diagnostics

-   parameteres package to visualise model parameters

## Visual statistical analysis with ggstatsplot

ggstats plot is ggplot2 extension to create a more detail information rich plot graphic

it is an alternative statistical inference method

it is a best practice for statistical reporting

## Getting Started

### Loading and installing package

```{r}
pacman::p_load(ggstatsplot, tidyverse)
```

### immporting data

```{r}
exam <- read_csv('data/Exam_data.csv')
```

## visualising ggstatsplot

### gghistostats: one-sample test

```{r}
set.seed(1234)

gghistostats(
  data = exam,
  x = ENGLISH,
  type = "bayes",
  test.value = 60,
  xlab = "English scores"
)
```

### ggbetweenstats: two-sample test

```{r}
ggbetweenstats(
  data = exam,
  x = GENDER, 
  y = MATHS,
  type = "np",
  messages = FALSE
)
```

### ggbetweenstats: one way ANOVE

```{r}
ggbetweenstats(
  data = exam,
  x = RACE, 
  y = ENGLISH,
  type = "p",
  mean.ci = TRUE, 
  pairwise.comparisons = TRUE, 
  pairwise.display = "s",
  p.adjust.method = "fdr",
  messages = FALSE
)
```

### ggscatterstats: Correlation significance test

```{r}
ggscatterstats(
  data = exam,
  x = MATHS,
  y = ENGLISH,
  marginal = FALSE,
  )
```

### ggbarstats: association significance test

```{r}
exam1 <- exam %>% 
  mutate(MATHS_bins = 
           cut(MATHS, 
               breaks = c(0,60,75,85,100))
)
```

```{r}
ggbarstats(exam1, 
           x = MATHS_bins, 
           y = GENDER)
```
